A tool for designing digital filters for the Hankel and Fourier transforms in potential, diffusive, and wavefield modeling

### 2018-11-22 ManuscriptGeophysics2018.tex
Final Submission

% \documentclass[manuscript]{geophysics}
% \documentclass[paper,twocolumn,twoside]{geophysics}
% \documentclass[paper]{geophysics}
% \documentclass[paper, revised]{geophysics}
\documentclass[manuscript,revised]{geophysics}

% Additional packages to geophysics.cls
\usepackage[USenglish]{babel}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage{amssymb, amsmath, amsfonts}
\usepackage{upquote}
\usepackage[strings]{underscore}
% \usepackage[pdftex, final]{hyperref}
% \hypersetup{allcolors=blue, allbordercolors={0 0 .5}, colorlinks=true}
\usepackage[pdftex, hidelinks]{hyperref}

@@

\address{
\footnotemark[1]TU Delft,
Building 23,
Stevinweg 1 / PO-box 5048,
2628 CN Delft,
E-mail: \href{mailto:D.Werthmuller@tudelft.nl}{D.Werthmuller@tudelft.nl};
\href{mailto:E.C.Slob@tudelft.nl}{E.C.Slob@tudelft.nl};
\footnotemark[2]Instituto Mexicano del Petróleo,
Eje Central Lázaro Cárdenas Norte 152,
Col. San Bartolo Atepehuacan C.P. 07730,
Ciudad de México, México,
\footnotemark[3]Lamont-Doherty Earth Observatory, Columbia University,
305C Oceanography,
61 Route 9W, PO Box 1000,
Palisades NY 10964-8000 US,
E-mail: \href{mailto:KKey@ldeo.columbia.edu}{KKey@ldeo.columbia.edu}.
}

@@

\cite{PhD.70.Ghosh} derived the filter coefficients in the spectral domain by
dividing the output spectrum by the input spectrum followed by an inverse
Fourier transform. Improvements to the determination of filter coefficients
were provided by \cite{EXG.75.ONeill, GEO.77.Nyman, GEO.82.Das}, or
specifically for the Fourier transform by \cite{GP.86.Nissen}. A direct
integration method was used by \cite{GP.76.Bichara} and \cite{GP.78.Bernabini}.
\cite{GP.79.Koefoed} proposed a Wiener-Hopf least-squares method, which was
further improved by many authors \citep{GP.82.Guptasarma, GEO.1982.Murakami,
GP.97.Guptasarma}. \cite{GP.07.Kong} proposed a direct matrix inversion method
to solve the convolution equation, which requires only the input and output
sample values. To evaluate the filter's effectiveness, he defines a good filter
as one that recovers small or weak diffusive EM fields. This method was also
used by \cite{GEO.09.Key, GEO.12.Key} to create a suite of filters accurate for
marine EM data. Most works have published filters for the Hankel transform with
$J_0$ and $J_1$ Bessel functions (or $J_{-1/2}$, $J_{1/2}$ if applied to the
Fourier sine/cosine transform), as all higher Bessel functions can be
rewritten, via recurrence relations, using only\old{use} these two.
\cite{GP.94.Mohsen} is one of the rare cases which provides $J_2$ filter
weights.

@@

All mentioned publications have in common that they were derived for direct
current methods (DC) or low frequency methods, such as time-domain shallow EM
methods (TEM), or controlled-source electromagnetics (CSEM), but not for high
frequency methods such as ground-pe\-ne\-tra\-ting radar (GPR). Generally it
was even thought that the filter method works only in the diffusive limit where
the quasistatic approximation is valid \citep[e.g.,][]{GEO.15.Hunziker}.
Nevertheless, \cite{GEO.17.Werthmuller} tested DLF for modeling 250 MHz
center-frequency GPR data using a 401 \old{point}\new{pt} filter that was
designed for diffusive EM, and found that it was orders of magnitude faster
than quadrature approaches. However, good quality results were only obtained
for the first arrivals at short time-offsets and the later arrivals had poor
quality. Yet, this promising initial test suggests it might be possible to
specifically design a special filter for GPR frequencies where the EM fields
propagate as waves rather than by diffusion.

@@

\subsection{Minimization criteria}
Figure~\ref{fig:Figure-1-bw} shows the differences between the different
minimization approaches; in (a) for a conventional, fast decaying transform
pair (using the $J_0$ kernel of a fullspace with vertical distance between
source and receiver of 50\,m, frequency $f=1\,$Hz, and resistivity
$\rho=1\,\Omega\,$m); and in (b) for a more complex, high frequency layered
model (using the $J_0$ kernel for $f=200\,$MHz for the GPR model described
in the section \emph{Ground-penetrating radar}, with receiver located at 1\,m
depth). The circles show the minimum amplitude used in previous approaches.
This criterion is not ideal, as it is subject to some random fluctuations and
depends on the choice of $r$. The squares show the minimum amplitude given a
certain acceptable error, and the diamonds show the maximum $r$ given a certain
acceptable error (the inversion is a minimization process, it therefore
minimizes $1/r$, not $r$). In simple cases the minimum amplitude given an
acceptable error and the maximum $r$ given an acceptable error will yield the
same result. However, in complex cases the maximum $r$ is more consistent and
therefore a much better criterion.
%
\plot*{Figure-1-bw}{width=.9\textwidth}{(a) A regular \new{right-hand-side
  (}RHS\new{)} curve of a transform pair with a purely decaying function.
  Minimum amplitude or maximum $r$ yield the same result in this scenario.
  Using a relative error criteria is more stable than just the absolute minimum
  amplitude. (b) A 1D model for $f=200\,$MHz. Here, the maximum $r$ provides a
  better criteria for the inversion.}
%

@@

Figure~\ref{fig:Figure-3-bw} shows four consecutive inversion runs for $N=201$,
where each run is a focus on a subsection of the previous run, indicated by the
\old{red} square. From the low resolution overview runs (a) and (b) it looks
like this is a standard, straight-forward minimization problem. However, from
the more detailed results in (c) and (d) it becomes obvious that it is a
minimization problem that has to be solved stochastically, as there are
solutions with 2 orders of magnitude difference apparently randomly next to
each other.
%
\plot*{Figure-3-bw}{width=\textwidth}{%
  Solution spaces of four consecutive inversion runs for a 201\,pt filter. Each
  consecutive run zooms into a portion of the previous solution space,
  indicated by the \old{red} square. The more in detail we obtain the solution
  space, the more random appears to be the distribution.}%
%

Figure~\ref{fig:Figure-4-bw} shows in (a) the filter values for $J_0$ and $J_1$
of the best 201 \old{point}\new{pt} filter obtained this way, and in (b) its
right-hand-side solution. The black dots indicate negative values, from which
it can \new{be} seen that adjacent values are often alternating between
positive and negative contributions.
%
\plot*{Figure-4-bw}{width=.9\textwidth}{(a) Filter values of the best obtained
  201\,pt filter with the corresponding check of quality in (b). Black points
  indicate negative values, which shows that adjacent values have often
  opposite signs.}
%

@@

\subsection{CSEM}
In this section we compare the 201\,pt filter derived in the previous section
to marine CSEM models \new{as} used in \cite{GP.07.Kong}, \new{in}
\cite{GEO.12.Key}, \old{as well as}\new{and} for a land EM case.

@@

Figure~\ref{fig:Figure-6-bw}(a) considers the canonical CSEM model as used in
\cite{GEO.12.Key} in his Figure~5: a $0.303\,\Omega\,$m sea with depth of
2\,km, a background resistivity of $1\,\Omega\,$m, in which a $100\,\Omega\,$m
target layer of 100\,m thickness is embedded at 1000\,m below the seafloor.
The source is located at 1990\,m and receivers are positioned on the seafloor.
%
\plot*{Figure-6-bw}{width=\textwidth}{(a) The canonical marine CSEM model of
  \cite{GEO.12.Key} and (b) a land model with source and receiver at the
  surface. \old{The new filter has generally relative errors which are several
  orders of magnitude lower than the other filters.} \new{The new filter shows
  relative errors that are generally several orders of magnitude lower than
  for the other filters.}
}
%

@@

It is very important to note again that other scenarios might yield very
different error plots. Although the new 201\,pt filter proves to be very
accurate for these three models, it might not be the best filter in other
cases. To verify the applicability of the filter we run two different tests
over a wide range of scenarios. Figure~\ref{fig:Figure-7} shows the relative
errors of DLF using the filter \emph{Wer201} compared to QWE for a deep water
model \new{(2\,km water depth)}, a shallow water model \new{(400\,m water
depth)}, and a land model. The models consist of air (upper halfspace), a water
layer in the first two examples, and a subsurface halfspace. Source depth is
1990\,m, 10\,m, and 0.5\,m in the deep, shallow, and land case, respectively.
Receiver depth is 2000\,m, 100\,m, and 0.8\,m. The subsurface halfspace
resistivity varies from 1 to 10, 100, and 1000\,$\Omega\,$m from the left to
the right column. The error plots cover offsets from 50\,m to 20\,km on the
x-axis, and 0.01\,Hz to 10\,Hz on the y-axis. It can be seen in the result that
the filter is sufficiently precise for all practical CSEM cases, hence has
generally an error far below 1\,\% ($10^{-2}$) in all cases. A relative noise
of less than 1\,\% is generally considered really good for real data. The
exceptions are in the shallow water case at large offsets with relatively high
frequencies, e.g., 10\,km offset with 8\,Hz; this zone has a very high error
for subsurface resistivities of 1\,$\Omega\,$m and 10\,$\Omega\,$m (yellow
zones). However, this is due to the fact that the amplitudes of the signal for
these offsets and frequencies are in the order of $10^{-25}\,$V/m. The grey
contour lines in the plot indicate the power of the amplitude of the signal.
%
\plot*{Figure-7}{width=\textwidth}{Relative error plots comparing the DLF
  (\emph{Wer201}) with QWE for the three cases deep water (top row), shallow
  water (middle row), and land (bottom row). The subsurface halfspace
  resistivity varies from 1 to 1000\,$\Omega\,$m from the left column to the
  right column; offsets vary from 50\,m to 20\,km, and frequencies vary from
  0.01\,Hz to 10\,Hz. The DLF provides sufficiently accurate results for all
  practical CSEM applications. The exception is in the shallow water case for
  $\rho = 1\,\Omega\,$m and $10\,\Omega\,$m for large offsets and high
  frequencies (yellow zones). However, the amplitudes in these regions are in
  the order of $10^{-25}\,$V/m, and therefore several orders of
  magnitude below the noise levels of current EM instrument systems.
  Contour lines show the power of the amplitude of the signal.}
%

@@

Figure~\ref{fig:Figure-9} shows the GPR example as calculated in
\cite{GEO.15.Hunziker} and \cite{GEO.17.Werthmuller}; the model parameters are
given in subplot (a). The filter used for this example for the Hankel transform
is a 2001\,pt filter, derived with the fullspace solution with $f=500\,$MHz for
the inversion and the check of quality. The whole space solution used
resistivity $200\,\Omega\,$m, relative permittivity $\varepsilon_\mr{r}=10$,
and permeability $\mu_\mr{r}=1$. Using Equation~\ref{eq:skin} yields a skin
depth of roughly 3.6\,m for these parameters. The source and receiver vertical
separation is 1\,m. For the Fourier transform a 4096\,pt FFT was used with
regularly spaced frequencies from 0.5\,MHz to 850\,MHz and then zero-padded up
to 2048\,MHz. The frequency result is multiplied with a Ricker wavelet with a
center frequency of 250\,MHz, and a gain function ($1 + |t^3|, t$ in ns) is
applied. Subplot (b) shows the result when using the adaptive quadrature (QUAD)
option of \texttt{empymod}, which uses the QAGSE routine from the Fortran
QUADPACK library. Subplot (c) shows the result using DLF with the new 2001\,pt
digital filter. The calculation with DLF took under 9 minutes and is therefore
roughly 80 times faster than the calculation with QUAD, which took roughly 11
hours and 27 minutes. Calculating the same model with QWE took 7 hours and 20
minutes. However, QWE uses in roughly 1/3 of the calculation internally QUAD in
this example, see \cite{GEO.17.Werthmuller}. Note that DLF was run in parallel
using 4 threads at once, taking effectively only 2 minutes and 10 seconds to
calculate.
% EMmod \citep{GEO.15.Hunziker} took for the same model about 18 hours 55
% minutes, using a 61\,pt Gauss-Kronrod integration routine.
The lagged convolution version of DLF and the splined version of QWE were used
in this comparison.
%
\plot*{Figure-9}{width=\textwidth}{GPR example for the model given in (a) using
  \texttt{empymod} with (b) \old{the} adaptive quadrature (QUAD) or with (c) a
  2001\,pt digital linear filter\old{ (b)}. QUAD took about 11\,h 27\,min to
  calculate, whereas DLF took 2\,min 10\,s in parallel on 4 nodes, hence under
  9\,min in total.}
%

Figure~\ref{fig:Figure-10-bw} shows in (a)-(c) the real \new{and imaginary}
part\new{s} of the frequency domain results and in (d)-(f) the time domain
results for offsets of 0.2\,m, 2.0\,m, and 3.0\,m.
%
\plot*{Figure-10-bw}{width=\textwidth}{The real\old{-part} \new{and imaginary
  parts of the} frequency-domain \new{(a-c)} and the time-domain \new{(d-f)}
  responses for \new{horizontal} offsets of (a/d) 0.2\,m, (b/e) 2.0\,m, and
  (c/f) 3.0\,m.}
%

The 2001\,pt filter was derived with the fullspace solution of a medium which
is very similar to the layer in which source and receiver reside in our GPR
example. To show that this filter can also be applied to different layer
parameters we run a test where we swapped layers one and two of the model in
Figure~\ref{fig:Figure-9} (a). The resulting real and imaginary parts for the
offsets 0.2\,m, 2.0\,m, and 3.0\,m are given in Figure~\ref{fig:Figure-11-bw}.
%
\plot*{Figure-11-bw}{width=\textwidth}{The real and imaginary parts of the
  frequency-domain response for the model given in Figure~\ref{fig:Figure-9}
  (a) with swapped first and second layers for \new{horizontal} offsets of (a)
  0.2\,m, (b) 2.0\,m, and (c) 3.0\,m.}
%

@@

The presented, free and open-source code \texttt{fdesign} can be used to design
digital linear filters for the Hankel and Fourier transforms (and more
generally for any linear transform) using either analytical transform pairs or
1D subsurface models together with the EM-modeler \texttt{empymod}. The code is
available from GitHub as part of \texttt{empymod}; the version corresponding to
this publication is \texttt{\new{v1.8.1}\old{v1.7.1.}}.

@@

D.W. would like to thank the electromagnetic research group of Aleksandr
Mousatov at the IMP for fruitful discussions, and Jürg Hunziker for valuable
feedback which improved this manuscript considerably. We thank the reviewers
for valuable inputs not only regarding the manuscript, but also regarding the
code and its documentation.

@@

\bibitem[Bichara and Lakshmanan, 1976]{GP.76.Bichara}
Bichara, M., and J. Lakshmanan,  1976, Fast automatic processing of resistivity
soundings\old{*}: Geophysical Prospecting, {\bf 24}, 354--370.
\newblock (\href{https://doi.org/10.1111/j.1365-2478.1976.tb00932.x}{doi:
  10.1111/j.1365-2478.1976.tb00932.x}).

@@

\bibitem[Das and Verma, 1981b]{GXP.81.Das}
--------, 1981b, The versatility of digital linear filters used in computing
resistivity and \old{em}\new{EM} sounding curves: Geoexploration, {\bf 18},
297--310. \newblock (\href{https://doi.org/10.1016/0016-7142(81)90059-4}{doi:
10.1016/0016-7142(81)90059-4}).

>>>>

% \documentclass[manuscript]{geophysics}
\documentclass[paper,twocolumn,twoside]{geophysics}
% \documentclass[paper]{geophysics}
% \documentclass[paper, revised]{geophysics}
% \documentclass[manuscript,revised]{geophysics}

% Additional packages to geophysics.cls
\usepackage[USenglish]{babel}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage{amssymb, amsmath, amsfonts}
\usepackage{upquote}
\usepackage[strings]{underscore}
\usepackage[pdftex, final]{hyperref}
\hypersetup{allcolors=blue, allbordercolors={0 0 .5}, colorlinks=true}
% \usepackage[pdftex, hidelinks]{hyperref}

@@

\address{
\footnotemark[1]TU Delft,
Building 23,
Stevinweg 1 / PO-box 5048,
2628 CN Delft,
E-mail: \href{mailto:D.Werthmuller@tudelft.nl}{D.Werthmuller@tudelft.nl};
\href{mailto:E.C.Slob@tudelft.nl}{E.C.Slob@tudelft.nl};
\footnotemark[2]\emph{formerly at the} Instituto Mexicano del Petróleo,
Eje Central Lázaro Cárdenas Norte 152,
Col. San Bartolo Atepehuacan C.P. 07730,
Ciudad de México, México,
\footnotemark[3]Lamont-Doherty Earth Observatory, Columbia University,
305C Oceanography,
61 Route 9W, PO Box 1000,
Palisades NY 10964-8000 US,
E-mail: \href{mailto:KKey@ldeo.columbia.edu}{KKey@ldeo.columbia.edu}.
}

@@

\cite{PhD.70.Ghosh} derived the filter coefficients in the spectral domain by
dividing the output spectrum by the input spectrum followed by an inverse
Fourier transform. Improvements to the determination of filter coefficients
were provided by \cite{EXG.75.ONeill, GEO.77.Nyman, GEO.82.Das}, or
specifically for the Fourier transform by \cite{GP.86.Nissen}. A direct
integration method was used by \cite{GP.76.Bichara} and \cite{GP.78.Bernabini}.
\cite{GP.79.Koefoed} proposed a Wiener-Hopf least-squares method, which was
further improved by many authors \citep{GP.82.Guptasarma, GEO.1982.Murakami,
GP.97.Guptasarma}. \cite{GP.07.Kong} proposed a direct matrix inversion method
to solve the convolution equation, which requires only the input and output
sample values. To evaluate the filter's effectiveness, he defines a good filter
as one that recovers small or weak diffusive EM fields. This method was also
used by \cite{GEO.09.Key, GEO.12.Key} to create a suite of filters accurate for
marine EM data. Most works have published filters for the Hankel transform with
$J_0$ and $J_1$ Bessel functions (or $J_{-1/2}$, $J_{1/2}$ if applied to the
Fourier sine/cosine transform), as all higher Bessel functions can be
rewritten, via recurrence relations, using only these two. \cite{GP.94.Mohsen}
is one of the rare cases which provides $J_2$ filter weights.

@@

All mentioned publications have in common that they were derived for direct
current methods (DC) or low frequency methods, such as time-domain shallow EM
methods (TEM), or controlled-source electromagnetics (CSEM), but not for high
frequency methods such as ground-pe\-ne\-tra\-ting radar (GPR). Generally it
was even thought that the filter method works only in the diffusive limit where
the quasistatic approximation is valid \citep[e.g.,][]{GEO.15.Hunziker}.
Nevertheless, \cite{GEO.17.Werthmuller} tested DLF for modeling 250 MHz
center-frequency GPR data using a 401 pt filter that was designed for diffusive
EM, and found that it was orders of magnitude faster than quadrature
approaches. However, good quality results were only obtained for the first
arrivals at short time-offsets and the later arrivals had poor quality. Yet,
this promising initial test suggests it might be possible to specifically
design a special filter for GPR frequencies where the EM fields propagate as
waves rather than by diffusion.

@@

\subsection{Minimization criteria}
Figure~\ref{fig:Figure-1-bw} shows the differences between the different
minimization approaches; in (a) for a conventional, fast decaying transform
pair (using the $J_0$ kernel of a fullspace with vertical distance between
source and receiver of 50\,m, frequency $f=1\,$Hz, and resistivity
$\rho=1\,\Omega\,$m); and in (b) for a more complex, high frequency layered
model (using the $J_0$ kernel for $f=200\,$MHz for the GPR model described
in the section \emph{Ground-penetrating radar}, with receiver located at 1\,m
depth). The circles show the minimum amplitude used in previous approaches.
This criterion is not ideal, as it is subject to some random fluctuations and
depends on the choice of $r$. The squares show the minimum amplitude given a
certain acceptable error, and the diamonds show the maximum $r$ given a certain
acceptable error (the inversion is a minimization process, it therefore
minimizes $1/r$, not $r$). In simple cases the minimum amplitude given an
acceptable error and the maximum $r$ given an acceptable error will yield the
same result. However, in complex cases the maximum $r$ is more consistent and
therefore a much better criterion.
%
\plot*{Figure-1-bw}{width=.9\textwidth}{(a) A regular right-hand-side (RHS)
  curve of a transform pair with a purely decaying function. Minimum amplitude
  or maximum $r$ yield the same result in this scenario. Using a relative error
  criteria is more stable than just the absolute minimum amplitude. (b) A 1D
  model for $f=200\,$MHz. Here, the maximum $r$ provides a better criteria for
  the inversion.}
%

@@

Figure~\ref{fig:Figure-3-bw} shows four consecutive inversion runs for $N=201$,
where each run is a focus on a subsection of the previous run, indicated by the
square. From the low resolution overview runs (a) and (b) it looks like this is
a standard, straight-forward minimization problem. However, from the more
detailed results in (c) and (d) it becomes obvious that it is a minimization
problem that has to be solved stochastically, as there are solutions with 2
orders of magnitude difference apparently randomly next to each other.
%
\plot*{Figure-3-bw}{width=\textwidth}{%
  Solution spaces of four consecutive inversion runs for a 201\,pt filter. Each
  consecutive run zooms into a portion of the previous solution space,
  indicated by the square. The more in detail we obtain the solution space, the
  more random appears to be the distribution.}%
%

Figure~\ref{fig:Figure-4-bw} shows in (a) the filter values for $J_0$ and $J_1$
of the best 201 pt filter obtained this way, and in (b) its right-hand-side
solution. The black dots indicate negative values, from which it can be seen
that adjacent values are often alternating between positive and negative
contributions.
%
\plot*{Figure-4-bw}{width=.9\textwidth}{(a) Filter values of the best obtained
  201\,pt filter with the corresponding check of quality in (b). Black points
  indicate negative values, which shows that adjacent values have often
  opposite signs.}
%

@@

\subsection{CSEM}
In this section we compare the 201\,pt filter derived in the previous section
to marine CSEM models as used in \cite{GP.07.Kong}, in \cite{GEO.12.Key},
and for a land EM case.

@@

Figure~\ref{fig:Figure-6-bw}(a) considers the canonical CSEM model as used in
\cite{GEO.12.Key} in his Figure~5: a $0.303\,\Omega\,$m sea with depth of
2\,km, a background resistivity of $1\,\Omega\,$m, in which a $100\,\Omega\,$m
target layer of 100\,m thickness is embedded at 1000\,m below the seafloor.
The source is located at 1990\,m and receivers are positioned on the seafloor.
%
\plot*{Figure-6-bw}{width=\textwidth}{(a) The canonical marine CSEM model of
  \cite{GEO.12.Key} and (b) a land model with source and receiver at the
  surface. The new filter shows relative errors that are generally several
  orders of magnitude lower than for the other filters.
}
%

@@

It is very important to note again that other scenarios might yield very
different error plots. Although the new 201\,pt filter proves to be very
accurate for these three models, it might not be the best filter in other
cases. To verify the applicability of the filter we run two different tests
over a wide range of scenarios. Figure~\ref{fig:Figure-7} shows the relative
errors of DLF using the filter \emph{Wer201} compared to QWE for a deep water
model (2\,km water depth), a shallow water model (400\,m water depth), and a
land model. The models consist of air (upper halfspace), a water layer in the
first two examples, and a subsurface halfspace. Source depth is 1990\,m, 10\,m,
and 0.5\,m in the deep, shallow, and land case, respectively. Receiver depth is
2000\,m, 100\,m, and 0.8\,m. The subsurface halfspace resistivity varies from 1
to 10, 100, and 1000\,$\Omega\,$m from the left to the right column. The error
plots cover offsets from 50\,m to 20\,km on the x-axis, and 0.01\,Hz to 10\,Hz
on the y-axis. It can be seen in the result that the filter is sufficiently
precise for all practical CSEM cases, hence has generally an error far below
1\,\% ($10^{-2}$) in all cases. A relative noise of less than 1\,\% is
generally considered really good for real data. The exceptions are in the
shallow water case at large offsets with relatively high frequencies, e.g.,
10\,km offset with 8\,Hz; this zone has a very high error for subsurface
resistivities of 1\,$\Omega\,$m and 10\,$\Omega\,$m (yellow zones). However,
this is due to the fact that the amplitudes of the signal for these offsets and
frequencies are in the order of $10^{-25}\,$V/m. The grey contour lines in the
plot indicate the power of the amplitude of the signal.
%
\plot*{Figure-7}{width=\textwidth}{Relative error plots comparing the DLF
  (\emph{Wer201}) with QWE for the three cases deep water (top row), shallow
  water (middle row), and land (bottom row). The subsurface halfspace
  resistivity varies from 1 to 1000\,$\Omega\,$m from the left column to the
  right column; offsets vary from 50\,m to 20\,km, and frequencies vary from
  0.01\,Hz to 10\,Hz. The DLF provides sufficiently accurate results for all
  practical CSEM applications. The exception is in the shallow water case for
  $\rho = 1\,\Omega\,$m and $10\,\Omega\,$m for large offsets and high
  frequencies (yellow zones). However, the amplitudes in these regions are in
  the order of $10^{-25}\,$V/m, and therefore several orders of
  magnitude below the noise levels of current EM instrument systems.
  Contour lines show the power of the amplitude of the signal.}
%

@@

Figure~\ref{fig:Figure-9} shows the GPR example as calculated in
\cite{GEO.15.Hunziker} and \cite{GEO.17.Werthmuller}; the model parameters are
given in subplot (a). The filter used for this example for the Hankel transform
is a 2001\,pt filter, derived with the fullspace solution with $f=500\,$MHz for
the inversion and the check of quality. The whole space solution used
resistivity $200\,\Omega\,$m, relative permittivity $\varepsilon_\mr{r}=10$,
and permeability $\mu_\mr{r}=1$. Using Equation~\ref{eq:skin} yields a skin
depth of roughly 3.6\,m for these parameters. The source and receiver vertical
separation is 1\,m. For the Fourier transform a 4096\,pt FFT was used with
regularly spaced frequencies from 0.5\,MHz to 850\,MHz and then zero-padded up
to 2048\,MHz. The frequency result is multiplied with a Ricker wavelet with a
center frequency of 250\,MHz, and a gain function ($1 + |t^3|, t$ in ns) is
applied. Subplot (b) shows the result when using the adaptive quadrature (QUAD)
option of \texttt{empymod}, which uses the QAGSE routine from the Fortran
QUADPACK library. Subplot (c) shows the result using DLF with the new 2001\,pt
digital filter. The calculation with DLF took under 9 minutes and is therefore
roughly 80 times faster than the calculation with QUAD, which took roughly 11
hours and 27 minutes. Calculating the same model with QWE took 7 hours and 20
minutes. However, QWE uses in roughly 1/3 of the calculation internally QUAD in
this example, see \cite{GEO.17.Werthmuller}. Note that DLF was run in parallel
using 4 threads at once, taking effectively only 2 minutes and 10 seconds to
calculate.
% EMmod \citep{GEO.15.Hunziker} took for the same model about 18 hours 55
% minutes, using a 61\,pt Gauss-Kronrod integration routine.
The lagged convolution version of DLF and the splined version of QWE were used
in this comparison.
%
\plot*{Figure-9}{width=\textwidth}{GPR example for the model given in (a) using
  \texttt{empymod} with (b) adaptive quadrature (QUAD) or with (c) a
  2001\,pt digital linear filter. QUAD took about 11\,h 27\,min to calculate,
  whereas DLF took 2\,min 10\,s in parallel on 4 nodes, hence under 9\,min in
  total.}
%

Figure~\ref{fig:Figure-10-bw} shows in (a)-(c) the real and imaginary
parts of the frequency domain results and in (d)-(f) the time domain results
for offsets of 0.2\,m, 2.0\,m, and 3.0\,m.
%
\plot*{Figure-10-bw}{width=\textwidth}{The real and imaginary parts of the
  frequency-domain (a-c) and the time-domain (d-f) responses for
  horizontal offsets of (a/d) 0.2\,m, (b/e) 2.0\,m, and (c/f) 3.0\,m.}
%

The 2001\,pt filter was derived with the fullspace solution of a medium which
is very similar to the layer in which source and receiver reside in our GPR
example. To show that this filter can also be applied to different layer
parameters we run a test where we swapped layers one and two of the model in
Figure~\ref{fig:Figure-9} (a). The resulting real and imaginary parts for the
offsets 0.2\,m, 2.0\,m, and 3.0\,m are given in Figure~\ref{fig:Figure-11-bw}.
%
\plot*{Figure-11-bw}{width=\textwidth}{The real and imaginary parts of the
  frequency-domain response for the model given in Figure~\ref{fig:Figure-9}
  (a) with swapped first and second layers for horizontal offsets of (a)
  0.2\,m, (b) 2.0\,m, and (c) 3.0\,m.}
%

@@

The presented, free and open-source code \texttt{fdesign} can be used to design
digital linear filters for the Hankel and Fourier transforms (and more
generally for any linear transform) using either analytical transform pairs or
1D subsurface models together with the EM-modeler \texttt{empymod}. The code is
available from GitHub as part of \texttt{empymod}; the version corresponding to
this publication is \texttt{v1.8.1}.

@@

D.W. would like to thank the electromagnetic research group of A. Mousatov at
the IMP for fruitful discussions, and J. Hunziker for valuable feedback which
improved this manuscript considerably. We thank the assistant editor J. Etgen,
the associate editor J. Dellinger, and the reviewers C. Weiss and R. Mittet
for valuable inputs not only regarding the manuscript, but also regarding the
code and its documentation.

@@

\bibitem[Bichara and Lakshmanan, 1976]{GP.76.Bichara}
Bichara, M., and J. Lakshmanan,  1976, Fast automatic processing of resistivity
  soundings Geophysical Prospecting, {\bf 24}, 354--370.
\newblock (\href{https://doi.org/10.1111/j.1365-2478.1976.tb00932.x}{doi:
  10.1111/j.1365-2478.1976.tb00932.x}).

@@

\bibitem[Das and Verma, 1981b]{GXP.81.Das}
--------, 1981b, The versatility of digital linear filters used in computing
resistivity and EM sounding curves: Geoexploration, {\bf 18}, 297--310.
\newblock (\href{https://doi.org/10.1016/0016-7142(81)90059-4}{doi:
10.1016/0016-7142(81)90059-4}).
